An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-19 docker-paper/manuscript.Rmd
separate dir for manuscript

---
title: "An introduction to Docker for reproducible research, with examples from the R environment"
author: 
  - name: Carl Boettiger
    affiliation: 
      l1: Center for Stock Assessment Research, 
      l2: 110 Shaffer Rd, Santa Cruz, CA 95050, USA 
    email: cboettig(at)gmail.com
    footnote: Corresponding author
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
abstract: |
  As computational work becomes more and more integral to many aspects
  of scientific research, computational reproducibility has become an
  issue of increasing importance to computer systems researchers and
  domain scientists alike. Though computational reproducibility seems 
  more straight forward than replicating physical experiments, the 
  immense, rapidly growing and frequently changing nature of computer
  environments makes being able to reproduce work in a way that others
  can understand and extend it a serious challenge.  In this paper, I
  explore common reasons that code developed for a gien research project
  cannot be successfully executed or extended by subsequent researchers.
  I review current approaches to these issues, including virtual machines
  and workflow systems, and their limitations. I then examine how the
  popular emerging technology Docker combines several areas from systems
  research - suc has operating system virtualization, cross-platform
  portability, modular re-usable elements, versioning, and a 'DevOps'
  philosophy, to address these challenges.  I illustrate this with several
  examples of Docker use with a focus on the R statistical environment.

@@

```bash
docker export <container-name> > container.tar
```

@@

```
FROM ubuntu:latest
RUN apt-get update && apt-get -y install r-recommended
```

@@

```bash
docker build -t r-recommended .
```

@@

```
FROM username/r-recommended 
RUN apt-get update && apt-get -y install r-cran-teachingdemos
```

@@

The effectiveness of this approach for supporting reproducible research
nonetheless depends on how each of these features are adopted and 
implemented by individual researchers. I summarize a few of the choices

- Use docker containers during development

A key feature of the Docker approach is the ability to mimic as closely
as possible the current workflow and development practices of the user.
Code executing inside a container on a local machine can appear identical
to code running natively, but with the added benefit that one can simply
recreate or snapshot and share the entire computational environment with
a few simple commands.  This works best if researchers set up their computational
environment in a container from the outset of the project.

- Write Dockerfiles instead installing interactive sessions.

As we have noted already, Docker can be used in a purely interactive
manner to record and distribute changes to a computational environment.
However, the approach is most useful for reproducible research when
researchers begin by defining their environment explicitly in the DevOps
fashion by writing a Dockerfile.

- Adding tests or checks to the Dockerfile

Dockerfile commands need not be limited to installing software, but
can also include execution.  This can help verify that an image has 
build successfully with all the software necessary to run the 
research code of interest.

- Use and provide appropriate base images

Though Docker supports modular design, it remains up to the researchers
to take advantage of it. An appropriate workflow might involve one
Dockerfile that includes all the software dependencies a researcher
usually uses in the course of their development, which can then be
extended by separate docker images for particular projects.  Re-using
existing images reduces the effort required to set up an environment,
contributes to the standardization of computational environments within
a field, and best leverages the ability of Docker's distribution system
to download only differences.

- Share Docker images and Dockerfiles

The Docker Hub significantly reduces the barriers for making even
large images (which can exceed the file size limits of journals common
scientific data repositories such as Dryad and Figshare) readily available
to other researchers.

- Archive tarball snapshots

Despite similar semantics to git, Docker's versioning system works rather 
differently than version management of code.  Docker can roll back AUFS
layers that have been added to an image, but not revert to the earlier 
state of a particular layer.  In consequence, to preserve a bitwise identical
snapshot of a container used to generate a given set of results, it is
necessary to archive the image tarball itself -- one can not simply
rely on the docker history to recover an earlier state.

@@

Docker does not provide complete virtualization but relies on the 
Linux kernel provided by the host.  Systems research can provide insight
on what limitations to reproducibility this introduces [e.g.  @Harji_2013].

Docker is limited to 64 bit host machines, making it impossible to run on
older hardware.

On Mac and Windows machines Docker must still be run in a fully
virtualized environment. Though the boot2docker tool streamlines this
process, it remains to be seen if the performance and integration with
the host machine's OS is sufficiently seamless or creates a barrier to
adoption by users on of these systems.

It remains to be seen if Docker will be significantly adopted by any scientific community.

@@

<!-- Is this really necessary? -->

@@

CB acknowledges support from NSF grant DBI-1306697, and also from the
Sloan Foundation support through the rOpenSci project. CB also wishes
to thank Dirk Eddelbuettel, Rich FitzJohn, Yihui Xie, Titus Brown,
John Stanton-Geddes and many others for helpful discussions about
reproducibility, virtualization, and Docker that have helped shape
this manuscript.

>>>>

---
title: "An introduction to Docker for reproducible research, with examples from the R environment"
author: 
  - name: Carl Boettiger
    affiliation: 
      l1: Center for Stock Assessment Research, 
      l2: 110 Shaffer Rd, Santa Cruz, CA 95050, USA 
    email: cboettig(at)gmail.com
    footnote: Corresponding author
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
abstract: |
  As computational work becomes more and more integral to many aspects of
  scientific research, computational reproducibility has become an issue
  of increasing importance to computer systems researchers and domain
  scientists alike. Though computational reproducibility seems more
  straight forward than replicating physical experiments, the complex
  and rapidly changing nature of computer environments makes being able
  to reproduce and extend such work a serious challenge.  In this paper,
  I explore common reasons that code developed for one research project
  cannot be successfully executed or extended by subsequent researchers.
  I review current approaches to these issues, including virtual machines
  and workflow systems, and their limitations. I then examine how the
  popular emerging technology Docker combines several areas from systems
  research - such as operating system virtualization, cross-platform
  portability, modular re-usable elements, versioning, and a 'DevOps'
  philosophy, to address these challenges.  I illustrate this with several
  examples of Docker use with a focus on the R statistical environment.

@@

```bash
docker export container-name > container.tar
```

@@

```
FROM ubuntu:latest
RUN apt-get update 
RUN apt-get -y install r-recommended
```

@@

```bash
docker build -t username/r-recommended .
```

@@

```
FROM username/r-recommended 
RUN apt-get update
RUN apt-get -y install r-cran-matrix
```

@@

The effectiveness of this approach for supporting reproducible research
nonetheless depends on how each of these features are adopted and 
implemented by individual researchers. I summarize a few of these practices here:

- _Use docker containers during development_. A key feature of the Docker
approach is the ability to mimic as closely as possible the current
workflow and development practices of the user.  Code executing inside
a container on a local machine can appear identical to code running
natively, but with the added benefit that one can simply recreate or
snapshot and share the entire computational environment with a few simple
commands.  This works best if researchers set up their computational
environment in a container from the outset of the project.

- _Write Dockerfiles instead installing interactive sessions_. As we
have noted already, Docker can be used in a purely interactive manner to
record and distribute changes to a computational environment.  However,
the approach is most useful for reproducible research when researchers
begin by defining their environment explicitly in the DevOps fashion by
writing a Dockerfile.

- _Adding tests or checks to the Dockerfile_. Dockerfile commands need
not be limited to installing software, but can also include execution.
This can help verify that an image has build successfully with all the
software necessary to run the research code of interest.

- _Use and provide appropriate base images_. Though Docker supports
modular design, it remains up to the researchers to take advantage of
it. An appropriate workflow might involve one Dockerfile that includes
all the software dependencies a researcher usually uses in the course of
their development, which can then be extended by separate docker images
for particular projects.  Re-using existing images reduces the effort
required to set up an environment, contributes to the standardization
of computational environments within a field, and best leverages the
ability of Docker's distribution system to download only differences.

- _Share Docker images and Dockerfiles_. The Docker Hub significantly
reduces the barriers for making even large images (which can exceed the
file size limits of journals common scientific data repositories such
as Dryad and Figshare) readily available to other researchers.

- _Archive tarball snapshots_. Despite similar semantics to git, Docker's
versioning system works rather differently than version management
of code.  Docker can roll back AUFS layers that have been added to
an image, but not revert to the earlier state of a particular layer.
In consequence, to preserve a bitwise identical snapshot of a container
used to generate a given set of results, it is necessary to archive the
image tarball itself -- one can not simply rely on the docker history
to recover an earlier state.

@@

- Docker does not provide complete virtualization but relies on the 
Linux kernel provided by the host.  Systems research can provide insight
on what limitations to reproducibility this introduces [e.g.  @Harji_2013].

- Docker is limited to 64 bit host machines, making it impossible to run on
older hardware.

- On Mac and Windows machines Docker must still be run in a fully
virtualized environment. Though the boot2docker tool streamlines this
process, it remains to be seen if the performance and integration with
the host machine's OS is sufficiently seamless or creates a barrier to
adoption by users on of these systems.

- Most importantly, it remains to be seen if Docker will be significantly
adopted by any scientific research or teaching community.

<!--

@@

<!-- Is this really necessary? Can it be more consise? -->

@@

CB acknowledges support from NSF grant DBI-1306697, and also from the
Sloan Foundation support through the rOpenSci project. CB also wishes
to thank Scott Chamberlain, Dirk Eddelbuettel, Rich FitzJohn, Yihui Xie,
Titus Brown, John Stanton-Geddes and many others for helpful discussions
about reproducibility, virtualization, and Docker that have helped shape
this manuscript.
